An agent-based model of military mechanization

### 2018-05-14 paper2/paper2.tex
foo

\begin{center}
	Word count: 4209
\end{center}
\section{Introduction}

@@

In summary, the orthodox view of military mechanization is that it is a
reflexive process: states choose their mechanization posture largely based on
that neighbors, enemies, and recent military experiences (e.g. defeat in
counter-insurgency). This paper argues
that this view---based on traditional longitudinal data analysis---is ripe for
agentization. It implicitly argues for autonomous agents (states) that have
cognition regarding mechanization that takes into account other agents (e.g.
enemies), spatiality (e.g. neighbors), and learning from experience. This paper
builds such a model and validates it against the empirical record, using the
National Mechanization Index of \cite{sechser2010army}. It finds that the 
orthodox view of mechanization does not stand up to agentization. A positive feedback loop exists causes states to
mechanize much faster and to a greater degree than occurs in the real-world.
This implies the existence of some countervailing factor (possibly a negative
feedback loop) that mitigates the extent of mechanization. Candidates for what
this could be mostly center on ``internal'' factors---e.g., culture, economy,
and so on---as opposed to the ``external'', security environment-centric
orthodox view.

@@

Importantly, the findings of \cite{sechser2010army} imply a positive feedback
loop. State A has high
mechanization, causing enemy State B to raise their mechanization, causing State
A to raise their mechanization... and so on. Such feedback loops are a hallmark
of complex systems and again are well-suited for representation by agent-based
modeling. This specific one---states changing their military in response to
other states change in military in response to... and so on---feedback loop---is often called
the ``security dilemma'' and was a main research interest of Thomas Schelling, one of the pioneers
of agent-based modeling (\cite{schelling1960strategy};
\cite{schelling2006micromotives}).

@@

I hypothesize that the arguments of \cite{sechser2010army} are incomplete and
that some omitted variable(s) act(s) as a ``brake'' on 
mechanization.\footnote{By ``omitted'', I mean that they conclude that the variable
is not very important for mechanization, not that they do not address the
variable at all.} An agent-based model is used to test this hypothesis. In broad
terms, agents (states) are defined to have the cognitive and learning behavior
implicitly claimed by \cite{sechser2010army}. The model is initialized with
real-world data from 1979 and allowed to run until 2001. The micro- and
macro-outcomes are then validated against real-world 2001 data. If
\cite{sechser2010army} are correct, the model's 2001 outcomes should match
real-world 2001 outcomes. If the model mechanization levels in 2001 are much
higher, then their claimed behavioral rules are incomplete and fail to include
some ``brake'' on mechanization, matching my hypothesis. Proving exactly what
that missing variable or set of variables is is outside the scope of this model
and paper, though it is discussed.

>>>>

\begin{center}
	Word count: 4218
\end{center}
\section{Introduction}

@@

In summary, the orthodox view of military mechanization is that it is a
reflexive process: states choose their mechanization posture largely based on
that neighbors, enemies, and recent military experiences (e.g. defeat in
counter-insurgency). This paper argues
that this view---based on traditional longitudinal data analysis---is ripe for
agentization. It implicitly argues for autonomous agents (states) that have
cognition regarding mechanization that takes into account other agents (e.g.
enemies), spatiality (e.g. neighbors), and learning from experience. This paper
builds such a model and validates it against the empirical record, using the
National Mechanization Index of \cite{sechser2010army}. It finds that the 
orthodox view of mechanization does not stand up to agentization. A positive feedback loop 
causes states to
mechanize much faster and to a greater degree than occurs in the real-world.
This implies the existence of some countervailing factor (possibly a negative
feedback loop) that mitigates the extent of mechanization. Candidates for what
this could be mostly center on ``internal'' factors---e.g., culture, economy,
and so on---as opposed to the ``external'', security environment-centric
orthodox view.

@@

Importantly, the findings of \cite{sechser2010army} imply a positive feedback
loop. State A has high
mechanization, causing enemy State B to raise their mechanization, causing State
A to raise their mechanization... and so on. Such feedback loops are a hallmark
of complex systems and again are well-suited for representation by agent-based
modeling. This specific feedback loop is often called
the ``security dilemma'' and was a main research interest of Thomas Schelling, one of the pioneers
of agent-based modeling (\cite{schelling1960strategy};
\cite{schelling2006micromotives}).

@@

I hypothesize that the arguments of \cite{sechser2010army} are incomplete and
that some omitted variable or set of variables acts as a ``brake'' on 
mechanization.\footnote{By ``omitted'', I mean that Sechser and Saunders conclude that the variable
is not very important for mechanization, not that they do not address the
variable at all.} An agent-based model is used to test this hypothesis. In broad
terms, agents (states) are defined to have the cognitive and learning behavior
implicitly claimed by \cite{sechser2010army}. The model is initialized with
real-world data from 1979 and allowed to run until 2001. The micro- and
macro-outcomes are then validated against real-world 2001 data. If
\cite{sechser2010army} are correct, the model's 2001 outcomes should match
real-world 2001 outcomes. If the model mechanization levels in 2001 are much
higher, then their claimed behavioral rules are incomplete and fail to include
some ``brake'' on mechanization, matching my hypothesis. Proving exactly what
that missing variable or set of variables is is outside the scope of this model
and paper, though it is discussed.

### 2018-05-14 paper2/prez.tex
foo

\documentclass{beamer}
\usetheme{Madrid}
\usecolortheme{spruce}

@@

@@

@@

	Sechser and Saunders 2010 conclude the former matters much more than the
	latter
	\begin{itemize}
		\item \small Developed state-year level dataset on
			mechanization, 1979-2001
		\item \small Traditional econometric analysis (OLS)
		\item \small Cultural norms (e.g. Irish tanks)\\~\\
	\end{itemize}

@@

\begin{frame}
	\frametitle{Model description}
	Agent = state actor\\~\\

	Attributes:
	\begin{itemize}
		\item \small Mechanization level (vehicles : infantry ratio,
			IISS data)
		\item \small Geographic neighbors (contiguous border per COW dataset)
		\item \small Enemies (defined based on 10-year MID history)
		\item \small Terrain (Fearon and Laitin 2004) \\~\\
	\end{itemize}

	Cognition regarding mechanization
	\begin{itemize}
		\item \small Takes into account neighbors, enemies, terrain,
			wars
		\item \small Agents observe lagged global behavior (2 years prior) and calculate
			weighted average of neighbor mechanization; weighted
			average of enemy mechanization; own terrain; and wars
		\item \small For each of the above, adjust own mechanization by some
			coefficient
		\item \small Coefficients drawn from statistical distribution
			found by Sechser and Saunders, randomly assigned to
			agents (heterogeneity)
	\end{itemize}
\end{frame}

\begin{frame}
	\frametitle{Model description}
	Initialized in 1979 \\~\\

	Proceeds in two-year timesteps until 2001 \\~\\

	Every turn, agents perceive neighbors, enemies, terrain, and wars, and
	adjust own mechanization accordingly \\~\\

@@

\begin{frame}
	\frametitle{Example}
\end{frame}

\begin{frame}
	\frametitle{Findings}

	Model results in significantly higher mechanization at both system- and
	agent-level than reality

	INSERT IMAGES
\end{frame}

\begin{frame}
	\frametitle{Influence of positive feedback loops}

	Results stem from positive feedback loop: you increase your
	mechanization, which causes me to increase my mechanization, which
	causes you to increase your mechanization... \\~\\

	Modeled system lacks countervailing negative feedback loop \\~\\

	Conclusion: some of the factors dismissed by Sechser and Saunders must
	play a role in mitigating rate-of-increase in the real-world system
	\\~\\

	Cannot say \textit{which} factors, but the system-level point remains

@@

\begin{frame}
	\frametitle{Verification, validation, and other topics}

	In our taxonomy: analyzed-analyzed model \\~\\

	Validation central to research strategy: comparison of model-2001 to
	reality-2001 \\~\\

	Significant amount of coding work was dealing with fact that composition
	of international system changed over time (e.g., 1991)

	\begin{itemize}
		\item \small Inheritance
	\end{itemize}
\end{frame}

\begin{frame}
	\frametitle{Possible extensions}

	Alliances \\~\\

	More sophisticated cognition \\~\\

	Incorporation of other force structure choices (e.g. human capital) \\~\\
\end{frame}

>>>>

\documentclass{beamer}
\usetheme{Madrid}

@@

\AtBeginSection[]{
	\begin{frame}
		\frametitle{Table of Contents}
		\tableofcontents[currentsection]
	\end{frame}
}

@@

\section{Background}

@@

	Sechser and Saunders 2010 conclude the former matters much more than the
	latter
	\begin{itemize}
		\item \small Developed state-year level dataset on
			mechanization, 1979-2001
		\item \small Traditional longitudinal data analysis (OLS)
	\end{itemize}

@@

\section{Model Description}

\begin{frame}
	\frametitle{Agents}
	Agent = state actor \\~\\

	Covers all countries in the world, minus microstates and ``inheritance'' problem

	\begin{figure}[h!]
		\centering
		\caption{Agents}
		\includegraphics[scale=0.04]{map.png}
	\end{figure}
\end{frame}

\begin{frame}
	\frametitle{Agent attributes}
	\begin{table}[h]
		\centering
		\begin{tabular}{|l l|}
			\hline
			\textbf{Identity} & \textit{Name} \\
			& \\
			\textbf{Mechanization} & \textit{Mechanization} \\
			& \\
			\textbf{Spatiality} & \textit{Neighbors} \\
			& \\
			\textbf{Inter-agent Relationships} & \textit{Enemies} \\
			& \\
			\textbf{Perception} & \textit{EnemiesMech} \\
			& \textit{NeighborsMech} \\
			& \textit{Lesson} \\
			& \\
			\textbf{Cognition} & \textit{EnemyMechSensitivity} \\
			& \textit{NeighborsMechSensitivity} \\
			& \textit{LessonSensitivity} \\
			\hline
		\end{tabular}
	\end{table}
\end{frame}

\begin{frame}
	\frametitle{Additional notes on agents and environment}
	Mechanization: combat vehicles per 100 infantry (data: IISS)\\~\\

	Spatiality: neighbor relationships rather than full-blown GIS \\~\\

	Enemies: defined on 10-year Militarized Interstate Dispute (MID) history\\~\\

	Perception: every time step, agents observe behavior of others and respond\\~\\

	Cognition: adjust own mechanization level based on neighbors, enemies, defeat in counter-insurgency
	\begin{itemize}
		\item \small Cognition is heterogenous across agents:
			coefficients drawn from distributions of Sechser and
			Saunders 
		\item \small Update equation:
			\begin{multline}
				\textrm{Mech}_i^{(t+1)} = \textrm{Mech}_i^{(t)}
				+ (\textrm{NeighborMechSensitivity}_i \times \textrm{NeighborMech}_i^{(t)}) \\
				+ (\textrm{EnemyMechSensitivity}_i \times \textrm{EnemyMech}_i^{(t)}) \\
				+ (\textrm{LessonSensitivity}_i \times I_{\textrm{Lesson}}(t, i))
			\end{multline}
	\end{itemize}
\end{frame}

\begin{frame}
	\frametitle{Time}
	Initialized in 1979, proceeds in two-year timesteps until 2001 \\~\\

	Every turn, agents (in random activation order):
	\begin{itemize}
		\item \small Perceive world
		\item \small Update assessment of neighbors, enemies, AML
		\item \small Adjust own mechanization \\~\\
	\end{itemize}

@@

\section{Findings}

\begin{frame}
	\frametitle{Model mechanization much higher at system-level than
	occurred in reality}
\begin{table}[h]
	\centering
	\caption{Comparison}
	\begin{tabular}{|l r r|}
		\hline
		\textbf{Year} & \textbf{Model} & \textbf{Reality} \\
		1979 & 0.0156 & 0.0139 \\
		1981 & 0.0189 & 0.0167 \\
		1983 & 0.0229 & 0.0167 \\
		1985 & 0.0280 & 0.0183 \\
		1987 & 0.0351 & 0.0187 \\
		1989 & 0.0415 & 0.0197 \\
		1991 & 0.0512 & 0.0215 \\
		1993 & 0.0636 & 0.0229 \\
		1995 & 0.0804 & 0.0231 \\
		1997 & 0.1014 & 0.0270 \\
		1999 & 0.1319 & 0.0272 \\
		2001 & 0.1730 & 0.0291 \\
		\hline
	\end{tabular}
\end{table}
\end{frame}

\begin{frame}
	\frametitle{Mechanization not uniform across agents - actually
	\textit{lower} in some}
	\begin{table}[h]
		\centering
		\caption{Mexico}
		\begin{tabular}{|l r r|}
			\hline
			\textbf{Year} & \textbf{Model} & \textbf{Reality} \\
			1979 & 0.0013 & 0.0012 \\
			1981 & 0.0014 & 0.0012 \\
			1983 & 0.0018 & 0.0026 \\
			1985 & 0.0019 & 0.0013 \\
			1987 & 0.0021 & 0.0019 \\
			1989 & 0.0022 & 0.0027 \\
			1991 & 0.0024 & 0.0030 \\
			1993 & 0.0026 & 0.0033 \\
			1995 & 0.0030 & 0.0035 \\
			1997 & 0.0032 & 0.0074 \\
			1999 & 0.0036 & 0.0082 \\
			2001 & 0.0040 & 0.0075 \\
			\hline
		\end{tabular}
	\end{table}
\end{frame}

\begin{frame}
	\frametitle{Findings driven by few outliers}
	\begin{table}[h]
		\centering
		\caption{Highest mechanization scores, 2001}
		\begin{tabular}{|l r|}
			\hline
			\textbf{Agent} & \textbf{Mechanization Score} \\
			Israel & 1.59 \\
			USA & 1.34 \\
			Kuwait & 1.14 \\
			Russia & 1.06 \\
			\hline
		\end{tabular}
	\end{table}
\end{frame}

\begin{frame}
	\frametitle{Outliers are clustered by neighborhood and relationship}
	\begin{table}[h]
		\centering
		\caption{Top 20 mechanized agents at simulation end, by region}
		\begin{tabular}{|l l l l l l|}
			\hline
			\textbf{MENA} & \textbf{Europe \& FSU} & \textbf{Africa} & \textbf{Americas} & \textbf{Asia} & \textbf{Oceania} \\
			Israel       & Russia         & Sudan        & USA    & (none) & (none) \\
			Kuwait       & France         & S. Africa    & Canada &        &        \\
			Iran         & UK             &              &        &        &        \\
			Libya        & Greece         &              &        &        &        \\
			Iraq         & Netherlands    &              &        &        &        \\
			Syria        & Poland         &              &        &        &        \\
			KSA          &                &              &        &        &        \\
			Egypt        &                &              &        &        &        \\
			Lebanon      &                &              &        &        &        \\
			Turkey       &                &              &        &        &        \\
			\hline
		\end{tabular}
	\end{table}
\end{frame}

\section{Discussion}

\begin{frame}
	\frametitle{Findings driven by positive feedback loop}

	State A raises mechanization, which causes neighbor / enemy State B to
	increase own mechanization... (\textit{security dilemma})\\~\\

	States with high exposure to this feedback loop due to many enemies and
	neighbors become outliers \\~\\

	Positive feedback loops are hallmark of complex system
	\begin{itemize}
		\item \small Autonomous, cognitive, heterogenous agents
		\item \small Spatiality
		\item \small Interaction
	\end{itemize}
\end{frame}

\begin{frame}
	\frametitle{Implications for larger literature}
	Findings imply orthodox, security environment-centric, ``external'' view
	is incomplete: there must be some variable or set of variables acting as
	a ``brake'' on military mechanization to mitigate the aforementioned
	positive feedback loop \\~\\

	Model cannot say exactly what those missing variables (but could assess
	whether candidates ``work'' in simulation...hint hint) \\~\\

	Agent-based models relatively rare in international relations (e.g.
	Cederman 1997, Axelrod 1997)
\end{frame}
\begin{frame}
	\frametitle{Larger course themes}

	Analyzed-analyzed\\~\\

	Empirical methods and validation central to research strategy \\~\\

	GIS in ABM: what level of detail is good enough? \\~\\

	Understanding behavior of complex models \\~\\

@@
